### 2015-05-31 chapters/encoding.tex
edits, bib, adherence to lang sci press formatting, etc.

\section{Encoding}
\label{encoding}

There are many in-depth histories of the origin and development of writing systems (e.g.~Robinson 1997, Powell 2012), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarised in this section, we have used mostly rather ephemeral internet sources. When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here. A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is Huurdeman (2003). Most books that discuss the development of encoding of telegraphic communication focus of cryptography (e.g.~Singh 1999) and forego the rather interesting story of `open' encoding that is related here.}

Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space. Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} `distant writing') have a long and widespread history since ancient times. The most widespread telegraphic systems worldwide are so-called whistled languages (Meyer 2015), but also drumming languages (Meyer et al. 2012) and smoke, fire, hydraulic or flag signals are forms of telegraphy.

Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances. The ``innovation'' of those systems was that all characters of the written language were replaced one-to-one by visual signals. Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.} This of course implies that the usefulness of any such telegraphic encoding completely depends on the sometimes rather haphazard structure of orthographic systems.

In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire. Originally, ``bisignal'' codes were used, consisting of two different signals. For example, Carl Friedrich Gauss in 1833 used positive and negative current (Mania 2008:282). More famous and influential, Samuel Morse in 1836 used long and short pulses.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed length of signals per character. In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems. An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as ``self synchronization''.}

In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605. However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as ``5-bit''. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called ``shift'' character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

@@

Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC~8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages. For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was actually a $8+4=12$-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2.176$ possible characters.}

@@

More wide-ranging, various people in the 1980s started to develop true international code sets. In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65.536$ different possible characters). At the same time in Europe, the \textsc{International Organization for Standardization} (ISO) was working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2.147.483.648$ possible characters, but because of some technical restrictions only 679.477.248 were allowed).

Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1.114.112$ possible characters). The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112.956 different characters (Unicode Consortium 2014).

>>>>

\chapter{Encoding}
\label{encoding}

There are many in-depth histories of the origin and development of writing systems (e.g.~\citet{Robinson1997,Powell2012}), a story that we will not repeat here. However, the history of turning writing into computer readable code is not so often told, so we decided to offer a short survey of the major developments here. This history turns out to be intimately related to the history of telegraphic communication.\footnote{Because of the recent history as summarized in this section, we have used mostly rather ephemeral internet sources. When not references by traditional literature in the bibliography, we have specifically used \url{http://www.unicode.org/history/} and various Wikipedia pages for the information presented here. A useful survey of the historical development of the physical hardware of telegraphy and telecommunication is \citet{Huurdeman2003}. Most books that discuss the development of encoding of telegraphic communication focus of cryptography, e.g.~\citet{Singh1999}, and forego the rather interesting story of `open' encoding that is related here.}

Writing systems have existed for roughly 6000 years, allowing people to exchange messages through time and space. Additionally, to bridge large geographic distances, \textsc{telegraphic systems} of communication (from Greek \emph{τῆλε γράφειν} `distant writing') have a long and widespread history since ancient times. The most widespread telegraphic systems worldwide are so-called whistled languages \citep{Meyer2015}, but also drumming languages \citep{Meyer_etal2012} and smoke, fire, hydraulic or flag signals are forms of telegraphy.

Telegraphy was reinvigorated in the end of the eighteenth century through the introduction of semaphoric systems by Claude Chapelle (since then using flags, flashing lights, or various specially designed contraptions) to convey messages over large distances. The \textit{innovation} of those systems was that all characters of the written language were replaced one-to-one by visual signals. Since then, all telegraphic systems have taken this principle, namely that any language to be transmitted first has to be turned into some orthographic system, which subsequently is encoded for transmission by the sender, and then turned back into orthographic representation at the receiver side.\footnote{Sound and video-based telecommunication of course takes a different approach by ignoring the written version of language and directly encode sound waves or light patterns.} This of course implies that the usefulness of any such telegraphic encoding completely depends on the sometimes rather haphazard structure of orthographic systems.

In the nineteenth century, \textsc{electric telegraphy} lead to a new approach in which written language characters were encoded by signals sent through a copper wire. Originally, \textsc{bisignal} codes were used, consisting of two different signals. For example, Carl Friedrich Gauss in 1833 used positive and negative current \citep[282]{Mania2008}. More famous and influential, Samuel Morse in 1836 used long and short pulses.

@@

From those bisignal encodings, true \textsc{binary codes} developed with a fixed length of signals per character. In such systems only a single separator between signals is needed, because the separation between characters can be established by counting until a fixed number of signals.\footnote{Of course, no explicit separator is needed when the timing of the signals is known, which is the principle used in all modern telecommunication systems. An important modern consideration is also how to know where to start counting when you did not catch the start of a message, something that is known in Unicode as \textsc{self synchronization}.}

In the context of electric telegraphy, such a binary code system was first established by Émile Baudot in 1870, using a fixed combination of five signals for each written character.\footnote{True binary codes have a longer history, going at least back to the Baconian cipher devised by Francis Bacon in 1605. However, the proposal by Baudot was the quintessential proposal leading to all modern systems.} There are $2^5 = 32$ possible combination when using five binary signals; an encoding today designated as \textsc{5-bit}. These codes are sufficient for all Latin letters, but of course they do not suffice for all written symbols, including punctuation and digits. As a solution, the Baudot code uses a so-called \textsc{shift} character, which signifies that from that point onwards---until shifted back---a different encoding is used, allowing for yet another set of 32 codes. In effect, this means that the Baudot code, and the \textsc{International Telegraph Alphabet} (ITA) derived from it, had an extra `bit' of information, so the encoding is actually 6-bit (with $2^6 = 64$ different possible characters). For decades, this encoding was the standard for all telegraphy and it is still in limited use today.

@@

Because 256 characters were clearly not enough for the many different characters needed in the world's writing systems, the ISO/IEC~8859 standard in 1987 extended the MCS to include 16 different 8-bit code pages. For example, part 5 was used for Cyrillic characters, part 6 for Arabic, and part 7 for Greek.\footnote{In effect, because $16 = 2^4$, this means that ISO/IEC~8859 was actually a $8+4=12$-bit encoding, though with very many duplicates by design (e.g.~all ASCII codes were repeated in each 8-bit code page). To be precise, ISO/IEC~8859 used the 7-bit ASCII as the basis for each code page, and defined 16 different 7-bit extensions, leading to $(1+16)\cdot{2^7} = 2,176$ possible characters.}

@@

More wide-ranging, various people in the 1980s started to develop true international code sets. In the United States, a group of computer scientists formed the \textsc{Unicode Consortium}, proposing a 16-bit encoding in 1991 (with $2^{16} = 65,536$ different possible characters). At the same time in Europe, the \textsc{International Organization for Standardization} (ISO) was working on ISO~10646 to supplant the ISO/IEC~8859 standard. Their first draft of the \textsc{Universal Character Set} (UCS) in 1990 was 31-bit (with theoretically $2^{31} = 2,147,483,648$ possible characters, but because of some technical restrictions only 679,477,248 were allowed).

Since 1991, the Unicode Consortium and the ISO jointly develop the \textsc{unicode standard}, or ISO/IEC~10646, leading to the current system including the original 16-bit Unicode proposal as the \textsc{basic multilingual plane}, and 16 additional planes of 16-bit for further extensions (with in total $(1+16) \cdot 2^{16} = 1,114,112$ possible characters). The most recent version of the Unicode Standard (7.0) was published in June 2014 and it defines 112,956 different characters \citep{Unicode2014}.

### 2015-05-31 chapters/introduction.tex
edits, bib, adherence to lang sci press formatting, etc.

\section{Introduction}
\label{introduction}

Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures. They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms. In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing `right' from `wrong' writing in schools). However, there is also a large community of users who take as many liberties in their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be the `right' way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out. As pointed out by the proponents of the `official' rules, there are some homographs when leaving out the tone diacritics (Olúmúyìw 2013: 44). However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language. More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation. By showing that you `know the rules' of expressing yourself in writing, others will more easily accept you as a worthy participant in their group. And that just as well holds for obeying to the `official' rules when writing a job application, as for obeying to the `informal' rules when writing an SMS to classmates in school. The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use (Olúmúyìw 2013).

@@

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity (cf.~Steiner et al. 2010, List 2012a, 2012b, List \& Moran 2013, Moran \& Prokic 2013). In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document. Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question. For example, the specification that <ou> is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words. Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).

@@

Before we deal with these proposals, in the first part of this paper (Sections 2 through 5) we give an extended introduction to the notion of encoding (Section 2) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section 3). We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard. A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section 4). As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with Unicode (Section 5).

The second part of the paper (Sections 6 and 7) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists. First, we discuss the challenges of characterizing a writing system. To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section 6). Finally, we discuss practical issues with actual examples (Section 7). We provide reference implementation of our proposals in R and in Python, available as open-source libraries.

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise (International Phonetic Association 2005). Standard conventions are used for distinguishing between graphemic < >, phonemic / / and phonetic [ ] representations. For character descriptions, we follow the notational conventions of the Unicode Standard (Unicode Consortium 2014). Character names are represented in small capital letters (e.g.~\textsc{latin small letter schwa}) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.

>>>>

% \section{Introduction}
% \label{introduction}

Writing systems arise and develop in a complex mixture of cultural, technological and practical pressures. They tend to be highly conservative, in that people who have learned to read and write in a specific way (however impractical or tedious) are mostly unwilling to change their habits, e.g.~they tend to resist spelling reforms. In all literate societies there exists a strong socio-political mainstream that tries to force unification of writing (for example by strongly enforcing ``right'' from ``wrong'' writing in schools). However, there is also a large community of users who take as many liberties in their writing as they can get away with.

For example, the writing of tone diacritics in Yoruba is often proclaimed to be the right way to write, although many users of Yoruba writing seem to be perfectly fine with leaving them out. As pointed out by the proponents of the official rules, there are some homographs when leaving out the tone diacritics \citet[44]{Olumuyiw2013}. However, writing systems (and the languages they represent) are normally full of homophones, which is normally not a problem at all for speakers of the language. More importantly, writing is not just a purely functional tool, but just as importantly it is a mechanism to signal social affiliation. By showing that you \textit{know the rules} of expressing yourself in writing, others will more easily accept you as a worthy participant in their group. And that just as well holds for obeying to the official rules when writing a job application, as for obeying to the informal rules when writing an SMS to classmates in school. The case of Yoruba writing is an exemplary case, as even after more than a century of efforts to standardize the writing systems, there is still a wide range of variation in daily use \citet{Olumuyiw2013}.

@@

The necessity to computationally use and compare different orthographies most forcefully arises in the context of language comparison. Concretely, in our current research our goal is to develop quantitative methods for language comparison and historical analysis in order to investigate worldwide linguistic variation and to model the historical and areal processes that underlie linguistic diversity, cf.~\citet{Steiner_etal2011,List2012,List2012a,ListMoran2013,MoranProkic2013}. In this work, it is crucial to be able to flexibly process across numerous resources with different orthographies. In many cases even different resources on the `same' language use different orthographic conventions. Another orthographic challenge that we encounter regularly in our linguistic practice is electronic resources on a particular language that claim to follow a specific orthographic convention (often a resource-specific convention), but on closer inspection such resources are almost always not consistently encoded. Thus a second goal of our orthography profiles is to allow for an easy specification of orthographic conventions, and use such profiles to check consistency and to report errors to be corrected.

A central step in our proposed solution to this problem is the tailored grapheme separation of strings of symbols, a process we call \textsc{grapheme tokenization}. Basically, given some strings of symbols (e.g.~morphemes, words, sentences) in a specific source, our first processing step is to specify how these strings have to be separated into graphemes, considering the specific orthographic conventions used in a particular source document. Our experience is that such a graphemic tokenization can be performed without extensive in-depth knowledge about the phonetic and phonological details of the language in question. For example, the specification that $<$ou$>$ is a grapheme of English is a much easier task than to specify what exactly the phonetic values of this grapheme are in any specific occurrence in English words. Grapheme separation is a task that can be performed relatively reliably and with limited availability of time and resources (compare, for example, the task of creating a complete phonetic or phonological normalization).

@@

Before we deal with these proposals, in the first part of this paper (Sections \ref{encoding} through \ref{ipa-meets-unicode}) we give an extended introduction to the notion of encoding (Section \ref{encoding}) and writing systems, both from a linguistic perspective and from the perspective of the Unicode Consortium (Section \ref{terminology}). We consider the Unicode Standard to be a breakthrough (and ongoing) development that fundamentally changed the way we look at writing systems, and we aim to provide here a slightly more in-depth survey of the many techniques that are available in the standard. A good appreciation for the solutions that the Unicode Standard also allows for a thorough understanding of possible pitfalls that one might encounter when using it (Section \ref{unicode-pitfalls}). As an example of the current state-of-the-art, we discuss the rather problematic marriage of the International Phonetic Alphabet (IPA) with the Unicode Standard (Section \ref{ipa-meets-unicode}).

The second part of the paper (Sections \ref{orthography-profiles} and \ref{use-cases}) describes our proposals for how to deal with the Unicode Standard in the daily practice of (comparative) linguists. First, we discuss the challenges of characterizing a writing system. To solve these problems, we propose the notions of orthography profiles, closely related to Unicode locale descriptions (Section \ref{orthography-profiles}). Finally, we discuss practical issues with actual examples (Section \ref{use-cases}). We provide reference implementation of our proposals in R and in Python, available as open-source libraries.

The following conventions are followed in this paper. All phonemic and phonetic representations are given in the International Phonetic Alphabet (IPA), unless noted otherwise \citep{IPA2005}. Standard conventions are used for distinguishing between graphemic < >, phonemic / / and phonetic [ ] representations. For character descriptions, we follow the notational conventions of the Unicode Standard \citep{Unicode2014}. Character names are represented in small capital letters (e.g.~\textsc{latin small letter schwa}) and code points are expressed as U\emph{+n} where \emph{n} is a four to six digit hexadecimal number, e.g.~U+0256, which can be rendered as the glyph <ə>.
